Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-22 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: reading through intro; acronyms; new refs

The large number of GW observations has allowed to infer relevant astrophysical~\cite{LIGOScientific:2021aug} and cosmological~\cite{LIGOScientific:2021psn} information on the compact-object population in our local Universe, and also to probe General Relativity (GR) in the high-velocity, dynamical and strong-field regime of gravity~\cite{LIGOScientific:2021sio}. The latter complement tests of GR in the low-velocity, quasi-static or linear regimes available with Solar-System experiments~\cite{Will:2014kxa}, binary-pulsar~\cite{Wex:2014nva} and galactic center~\cite{GRAVITY:2018ofz,Do:2019txf,EventHorizonTelescope:2019ths} observations, 
and cosmological measurements~\cite{Clifton:2011jh}.

The coalescence of two BHs in GR is characterized by a long \textit{inspiral} stage, during which the 
holes adiabatically and steadly come closer and closer to each other, loosing energy because of the emission 
of GWs. Then, they \textit{merge}, forming a common apparent horizon. Subsequently, during the \textit{ringdown} stage, the newly formed remnant object settles down to a Kerr BH emitting quasi-normal modes (QNMs)~\cite{Vishveshwara:1970cc,Press:1971wr,Chandrasekhar:1975zza}. Because of the no-hair conjecture in GR~\cite{Carter:1971zc,Israel:1967wq,Hawking:1971vc,Robinson:1975bv}, the QNM (complex) frequencies of (electrically neutral) astrophysical BHs are only described by the BH's mass and spin. In GR, the QNM frequencies are labeled by the harmonic indices $(\ell,m)$ and the overtone number $n$.

Several null tests have been proposed to probe the nature of gravity with GW signals~\cite{TheLIGOScientific:2016src,Yunes:2016jcc,Abbott:2018lct,LIGOScientific:2019fpa,Abbott:2020jks,LIGOScientific:2021sio}. They include tests of GW generation~\cite{Arun:2006yw,Yunes:2009ke,Li:2011cg,Agathos:2013upa,Mehta:2022pcn}, where deviations in the post-Newtonian (PN) coefficients in the inspiral, and phenomenological coefficients in the intermediate (plunge) and merger stages can be bounded; tests of GW propagation~\cite{Will:1997bb}, which allow to set upper limit on coefficients entering generalized dispersion relations, including the Compton wavelength associated to the 
mass of the graviton; tests of the polarization of gravitational radiation~
\cite{Will:2014kxa}, for which more than 
two GW detectors are needed to set statistically significant bounds, and tests of the remnant properties~\cite{Meidam:2014jpa,Carullo:2018sfu,Brito:2018rfr,Carullo:2019flw,Isi:2019aib,Ghosh:2021mrv,Carullo:2021dui} 
during the post-merger stage. So far, none of these null tests have reported any deviation from GR.

Probing the gravitational properties of the remnant object during the
ringdown, has attracted a lot of attention in the last twenty years. 
Reference~\cite{Dreyer:2003bv} proposed the idea of employing spectroscopy
of the ringdown stage to rule out (or constrain) either modified theories of 
gravity or exotic compact objects (in GR) rather than BHs, thus 
testing the no-hair conjecture. Since then, many studies have quantified the accuracy 
with which the QNM (complex) frequencies can be measured for
GW sources detectable with ground- and space-based detectors (see, e.g., Refs.~\cite{Berti:2005ys,Baibhav:2020tma,Bhagwat:2021kwv}). Several analyses~\cite{Carullo:2018sfu,
Brito:2018rfr,Carullo:2019flw,Isi:2019aib,Ghosh:2021mrv,Carullo:2021dui}  
have used the GW observations of the LIGO-Virgo-KAGRA (LVK) collaboration to set upper 
limits on deviations in the QNM (complex) frequencies of BHs in GR. Others have claimed 
the measurements of QNMs beyond the dominant $(2,2,0)$ mode~\cite{Capano:2021etf}, 
or overtones --- for example the $(2,2,1)$ ymode~\cite{Isi:2019aib,Cotesta:2022pci,Isi:2022mhy}. These studies have been pursued either using a superposition 
of damped sinusoids~\cite{Giesler:2019uxc,Carullo:2019flw}, in some cases augmemted with QNM amplitudes calibrated to 
numerical-relativity simulations, or with parameterized inspiral-merger-ringdown (IMR) 
waveform models, where the QNM (complex) frequencies are not necessarily fixed 
to the GR values for BHs, but kept free~\cite{Brito:2018rfr,Ghosh:2021mrv}.

Here, we will employ the parameterized IMR model of Ref.~\cite{Ghosh:2021mrv}, constructed from a nonprecessing-spin effective-one-body (EOB) waveform model~\cite{Bohe:2016gbl,Cotesta:2018fcv,Mihaylov:2021bpf}, to carry out theory-specific tests of the 
ringdown using four high-curvature gravity theories. Previously, such parameterized 
waveform model was employed in Refs.~\cite{Abbott:2020jks,LIGOScientific:2021sio} for theory-independent tests of the ringdown. More specifically, here we will focus on four modified gravity theories, Einstein-dilaton Gauss-Bonnet gravity, dynamical 
Chern-Simon gravity, cubic and quartic effective-field theories of GR, and 
express the QNM (complex) frequencies using the parametrized ringdown spin-expansion coefficients (ParSpec) 
of Ref.~\cite{Maselli:2019mjd}. In such a framework, the non-GR QNM frequencies are  
recast as deviations from the GR QNM values, and are expressed in terms of a single-free parameter, 
the fundamental length-scale $\ell_{\rm th}$ of the gravity theory under consideration, the GR limit 
corresponding to $\ell_{\rm th} \rightarrow 0$.

With this formulation of the ringdown, we use the two loudest merger-ringdown GW events, so far 
observed by the LVK collaboration, notably GW150914 and GW200129, and Bayesian-inference techniques to 
perform null tests. We find that GR is not violated, and we set upper limits, at $90\%$ credible level, 
on the length-scale $\ell_{\rm th}$. In Table~\ref{tab:bound_summary}, we summarize our results, and compare 
them with existing constraints.

\begin{table}[th]
\begin{tabular}{c | c c}
\hline \hline
Theory & Constraint & This work \\
\hline
EdGB        & $\ell_{\rm GB} \leqslant 1.18$~km~(GW)~\cite{Lyu:2022gdr} & -- \\
dCS         & $\ell_{\rm CS} \leqslant 8.5$~km~(EM+GW)~\cite{Silva:2020acr}  & $\ell_{\rm CS} \leqslant 38.7$~km \\
cubic EFT   & -- & $\ell_{\rm cEFT} \leqslant 38.2$~km \\
quartic EFT & $\ell_{\rm qEFT} \leqslant 150$~km~\cite{Sennett:2019bpc}  & $\ell_{\rm qEFT} \leqslant 51.3$~km \\
\hline \hline
\end{tabular}
\caption{Summary of the upper limits, at $90\%$ credible level, 
on the length-scale of the modified gravity theories under investigation. 
They were obtained combining Bayesian-inference results from GW150914 and GW200129. 
Current constraints on $\ell_{\rm th}$ are also listed.}
\label{tab:bound_summary}
\end{table}

The paper is organized as follows. In Sec.~\ref{sec:review_theories}, we briefly describe the four 
higher-curvature modified gravity theories for which we peform the ringdown test. 
In Sec.~\ref{sec:method} we build our parameterized IMR model for nonprecessing-spin compact-object 
binaries making use of the ParSpec framework. After reviewing the Bayesian inference method, 
in Sec.~\ref{sec:pe}, we motivate our selection of GW events from the LVK catalog, and also 
discuss the range of validity of our analyses. In particular, we discuss the impact on 
our results of the assumptions underlying the ParSpec framework, and the fact that 
our modified gravity theories have to be interpreted as EFTs. In Sec.~\ref{sec:results}, 
we present our results obtained by applying Bayesian analysis on the LVK data of 
GW150914 and GW1200129, and discuss how we set the upper limits 
on the fundamental length-scales $\ell_{\rm th}$. Finally, in Sec.~\ref{sec:conclusions} we 
summarize our findings, and discuss how to make our framework more robust, in view 
of stronger GW events in upcoming GW observational runs, by including  physical effects currently absent in our 
study (e.g., precessing-spins and eccentricity). In the Appendix~\ref{app:map_details} 
we provide details in calculating the nonGR QNM (complex) frequencies, when using ParSpec,  
for the four modified gravity theories under consideration.

Henceforth, unless otherwise specified, we work in geometric units $G = 1 = c$.

@@

This theory belongs to the class of scalar-Gauss-Bonnet theories, which are
described by the action
%
\begin{equation} \label{eq:action_sgb}
    S_{\rm EdGB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm EdGB} f(\varphi) \, \mathcal{G}
    \right]\,,
\end{equation}
%
where $g \equiv \textrm{det}(g_{\mu \nu})$ is the metric determinant, $R$ is the Ricci
scalar, $\varphi$ is a dynamical scalar field that couples to the
Gauss-Bonnet invariant
%
$\mathcal{G} =
R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
- 4 R^{\mu\nu}R_{\mu\nu}
+ R^2$,
%
and $R_{\mu\nu\rho\sigma}$ and $R_{\mu\nu}$ are the Riemann and Ricci tensors respectively.
%
By itself, $\int \dV \mathcal{G}$ is a boundary term in four dimensions
and hence does not contribute to the field equations~\cite{Myers:1987yn}.
%
However,  when coupled to $\varphi$, it can contribute to the field equation
through the coupling function $f(\varphi)$. The strength of the coupling is
set by $\ell_{\rm GB}$, with dimensions of length.

@@

An accurate description of the merger is incorporated through calibration with
numerical relativity (NR) simulations, as described in Refs.~\cite{Bohe:2016gbl,Cotesta:2018fcv}, along with
information for the merger and ringdown phases, from BH perturbation theory.
The merger-ringdown waveform, $h_{l m}^\textrm{merger-RD}$, is then stitched to
inspiral-plunge waveform, $h_{l m}^\textrm{insp-plunge}$ at a certain time $t = t^{\textrm{match}}_{l m}$, as
%
\begin{equation}
h_{l m}(t) =
h_{l m}^\textrm{insp-plunge}\, \Theta(t_\textrm{match}^{l m} - t)
+ h_{l m}^\textrm{merger-RD}\,\Theta(t-t_\textrm{match}^{l m})\,,
\end{equation}
%
where $\Theta(t)$ is the Heaviside step function.
%
The merger-ringdown waveform is expressed as an exponentially damped
sinusoid~\cite{Bohe:2016gbl,Cotesta:2018fcv}
%
\begin{equation}
\label{RD}
h_{l m}^{\textrm{merger-RD}}(t) = \nu \ \tilde{A}_{l m}(t)\ e^{i \tilde{\phi}_{l m}(t)} \ e^{-i \sigma_{l m 0}(t-t^{\textrm{match}}_{l m})}\,,
\end{equation}
%
where
%
\begin{align}
\sigma_{l m 0} \equiv {\rm Re}(\sigma_{l m 0}) + i \, {\rm Im}(\sigma_{l m 0})
= \omega_{l m 0} - \frac{i}{\tau_{l m 0}}\,,
\end{align}
%
are the complex frequencies of the fundamental ($0$-th overtone) QNMs of the remnant BH.
%
%\ag{Going back to the convention established above to use the index $k$ to now denote the $(l, m,0)$
%mode,  the frequencies $\omega_k$ and damping times $\tau_k$ can be read off from the
%real and imaginary parts of $\sigma_k$ respectively.}
%
The functions $\tilde{A}_{l m}(t)$ and $\tilde{\phi}_{l m}(t)$ are defined
in Ref.~\cite{Bohe:2016gbl,Cotesta:2018fcv}.

>>>>

The large number of GW observations has allowed us to infer relevant astrophysical~\cite{LIGOScientific:2021aug} and cosmological~\cite{LIGOScientific:2021psn} information on the compact-object population in our local Universe, and also to probe General Relativity (GR) in the high-velocity, dynamical and strong-field regime of gravity~\cite{LIGOScientific:2021sio}. The latter complement tests of GR in the low-velocity, quasi-static or linear regimes available with Solar-System experiments~\cite{Will:2014kxa}, binary-pulsar~\cite{Wex:2014nva,Kramer:2021jcw} and galactic-center~\cite{GRAVITY:2018ofz,Do:2019txf,EventHorizonTelescope:2019ths} observations,
and cosmological measurements~\cite{Clifton:2011jh}.

The coalescence of two BHs in GR is characterized by a long \textit{inspiral} stage, during which the
holes adiabatically and steadly come closer and closer to each other, loosing energy because of the emission
of GWs. Then, they \textit{merge}, forming a common apparent horizon. Subsequently, during the \textit{ringdown} stage, the newly formed remnant object settles down to a Kerr BH emitting quasinormal modes (QNMs)~\cite{Vishveshwara:1970cc,Press:1971wr,Chandrasekhar:1975zza}. Because of the no-hair conjecture in GR~\cite{Carter:1971zc,Israel:1967wq,Hawking:1971vc,Robinson:1975bv}, the QNM (complex) frequencies of (electrically neutral) astrophysical BHs are only described by the BH's mass and spin. In GR, the QNM frequencies are labeled by the harmonic indices $(\ell,m)$ and the overtone number $n$.

Several null tests have been proposed to probe the nature of gravity with GW signals~\cite{TheLIGOScientific:2016src,Yunes:2016jcc,Abbott:2018lct,LIGOScientific:2019fpa,Abbott:2020jks,LIGOScientific:2021sio}. They include tests of GW generation~\cite{Arun:2006yw,Yunes:2009ke,Li:2011cg,Agathos:2013upa,Mehta:2022pcn}, where deviations in the post-Newtonian (PN) coefficients in the inspiral, and phenomenological coefficients in the intermediate (plunge) and merger stages can be bounded; tests of GW propagation~\cite{Will:1997bb}, which allow to set upper limit on coefficients entering generalized dispersion relations, including the Compton wavelength associated to the
mass of the graviton; tests of the polarization of gravitational radiation~\cite{Will:2014kxa}, for which more than
two GW detectors are needed to set statistically significant bounds, and tests of the remnant properties~\cite{Meidam:2014jpa,Carullo:2018sfu,Brito:2018rfr,Carullo:2019flw,Isi:2019aib,Ghosh:2021mrv,Carullo:2021dui}
during the post-merger stage. So far, none of these null tests have reported any deviation from GR.

Probing the gravitational properties of the remnant object during the
ringdown, has attracted a lot of attention in the last twenty years.
Reference~\cite{Dreyer:2003bv} proposed the idea of employing BH spectroscopy~\cite{Detweiler:1980gk}
of the ringdown stage to rule out (or constrain) either modified theories of
gravity or exotic compact objects (in GR) rather than BHs, thus
testing the no-hair conjecture. Since then, many studies have quantified the accuracy
with which the QNM frequencies can be measured for
GW sources detectable with ground- and space-based detectors (see, e.g., Refs.~\cite{Berti:2005ys,Baibhav:2020tma,Bhagwat:2021kwv,Ota:2021ypb}). Several analyses~\cite{Carullo:2018sfu,
Brito:2018rfr,Carullo:2019flw,Isi:2019aib,Ghosh:2021mrv,Carullo:2021dui}
have used the GW observations of the LIGO-Virgo-KAGRA (LVK) collaboration to set upper
limits on deviations in the QNM frequencies of BHs in GR. Others have claimed
the measurements of QNMs beyond the dominant $(2,2,0)$ mode~\cite{Capano:2021etf},
or overtones --- for example the $(2,2,1)$ mode~\cite{Isi:2019aib,Cotesta:2022pci,Isi:2022mhy}.
%
These studies have been pursued either using a superposition
of damped sinusoids~\cite{Giesler:2019uxc,Carullo:2019flw}, in some cases augmemted with QNM amplitudes calibrated to
numerical-relativity (NR) simulations, or with parameterized inspiral-merger-ringdown (IMR)
waveform models, where the QNM frequencies are not necessarily fixed
to the GR values for BHs, but kept free~\cite{Brito:2018rfr,Ghosh:2021mrv}.

Here, we will employ the parameterized IMR model of Ref.~\cite{Ghosh:2021mrv}, constructed from a nonprecessing-spin effective-one-body (EOB) waveform model~\cite{Bohe:2016gbl,Cotesta:2018fcv,Mihaylov:2021bpf}, to carry out theory-specific tests of the
ringdown using four high-curvature gravity theories. Previously, such parameterized
waveform model was employed in Refs.~\cite{Abbott:2020jks,LIGOScientific:2021sio} for theory-independent tests of the ringdown.
%
More specifically, here we will focus on four modified gravity theories, Einstein-dilaton-Gauss-Bonnet gravity, dynamical
Chern-Simon gravity, cubic and quartic effective-field theories (EFTs) of GR, and
express the QNM frequencies using the parametrized ringdown spin-expansion coefficients (ParSpec)
of Ref.~\cite{Maselli:2019mjd}.
%
In this framework, the non-GR QNM frequencies are
recast as deviations from the GR QNM values, and are expressed in terms of a single free parameter,
the fundamental length-scale $\ell_{\rm th}$ of the gravity theory under consideration, the GR limit
corresponding to $\ell_{\rm th} \rightarrow 0$.

With this formulation of the ringdown, we use the two loudest merger-ringdown GW events, so far
observed by the LVK collaboration, notably GW150914 and GW200129, and use Bayesian-inference techniques to
perform null tests. We find no indication that GR is violated and, when possible, we place upper limits, at $90\%$ credible level,
on the length-scale $\ell_{\rm th}$ of each theory. In Table~\ref{tab:bound_summary}, we summarize our results, and compare
them with existing constraints.

\begin{table}[t]
\begin{tabular}{c | c c}
\hline \hline
Theory & Constraint & This work \\
\hline
EdGB        & $\ell_{\rm EdGB} \leqslant 1.18$~km~(GW)~\cite{Lyu:2022gdr} & -- \\
dCS         & $\ell_{\rm dCS} \leqslant 8.5$~km~(EM+GW)~\cite{Silva:2020acr}  & $\ell_{\rm dCS} \leqslant 38.7$~km \\
cubic EFT   & -- & $\ell_{\rm cEFT} \leqslant 38.2$~km \\
quartic EFT & $\ell_{\rm qEFT} \leqslant 150$~km~\cite{Sennett:2019bpc}  & $\ell_{\rm qEFT} \leqslant 51.3$~km \\
\hline \hline
\end{tabular}
\caption{Summary of the upper limits, at $90\%$ credible level,
on the length-scale of the modified gravity theories under investigation.
They were obtained combining Bayesian-inference results from GW150914 and GW200129.
Current constraints on $\ell_{\rm th}$ are also listed.}
\label{tab:bound_summary}
\end{table}

The paper is organized as follows. In Sec.~\ref{sec:review_theories}, we briefly describe the four
higher-curvature modified gravity theories for which we peform the ringdown test.
In Sec.~\ref{sec:method} we build our parameterized IMR model for nonprecessing-spin compact-object
binaries making use of the ParSpec framework. After reviewing the Bayesian inference method,
in Sec.~\ref{sec:pe}, we motivate our selection of GW events from the LVK catalog, and also
discuss the range of validity of our analyses. In particular, we discuss the impact on
our results of the assumptions underlying the ParSpec framework, and the fact that
our modified gravity theories have to be interpreted as EFTs. In Sec.~\ref{sec:results},
we present our results obtained by applying Bayesian analysis on the LVK data of
GW150914 and GW1200129, and discuss how we set the upper limits
on the fundamental length-scales $\ell_{\rm th}$. Finally, in Sec.~\ref{sec:conclusions} we
summarize our findings, and discuss how to make our framework more robust, in view
of stronger GW events in upcoming GW observational runs, by including  physical effects currently absent in our
study (e.g., precessing-spins and eccentricity). In the Appendix~\ref{app:map_details}
we provide details in calculating the non-GR QNM frequencies, when using ParSpec,
for the four modified gravity theories under consideration.
%
Henceforth, unless otherwise specified, we work in geometric units $G = 1 = c$.

@@

This theory belongs to the class of scalar-Gauss-Bonnet theories, which are
described by the action
%
\begin{align} \label{eq:action_sgb}
    S_{\rm EdGB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm EdGB} f(\varphi) \, \mathcal{G}
    \right]\,,
    \nonumber \\
\end{align}
%
where $g \equiv \textrm{det}(g_{\mu \nu})$ is the metric determinant, $R$ is the Ricci
scalar, $\varphi$ is a dynamical scalar field that couples to the
Gauss-Bonnet invariant
%
$\mathcal{G} =
R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
- 4 R^{\mu\nu}R_{\mu\nu}
+ R^2$,
%
and $R_{\mu\nu\rho\sigma}$ and $R_{\mu\nu}$ are the Riemann and Ricci tensors respectively.
%
By itself, $\int \dV \mathcal{G}$ is a boundary term in four dimensions
and hence does not contribute to the field equations~\cite{Myers:1987yn}.
%
However,  when coupled to $\varphi$, it can contribute to the field equation
through the coupling function $f(\varphi)$. The strength of the coupling is
set by $\ell_{\rm GB}$, with dimensions of length.

@@

An accurate description of the merger is incorporated through calibration with
NR simulations, as described in Refs.~\cite{Bohe:2016gbl,Cotesta:2018fcv}, along with
information for the merger and ringdown phases, from BH perturbation theory.
The merger-ringdown waveform, $h_{l m}^\textrm{merger-RD}$, is then stitched to
inspiral-plunge waveform, $h_{l m}^\textrm{insp-plunge}$ at a certain time $t = t^{\textrm{match}}_{l m}$, as
%
\begin{equation}
h_{l m}(t) =
h_{l m}^\textrm{insp-plunge}\, \Theta(t_\textrm{match}^{l m} - t)
+ h_{l m}^\textrm{merger-RD}\,\Theta(t-t_\textrm{match}^{l m})\,,
\end{equation}
%
where $\Theta(t)$ is the Heaviside step function.
%
The merger-ringdown waveform is expressed as an exponentially damped
sinusoid~\cite{Bohe:2016gbl,Cotesta:2018fcv}
%
\begin{equation}
\label{RD}
h_{l m}^{\textrm{merger-RD}}(t) = \nu \ \tilde{A}_{l m}(t)\ e^{i \tilde{\phi}_{l m}(t)} \ e^{-i \sigma_{l m 0}(t-t^{\textrm{match}}_{l m})}\,,
\end{equation}
%
where
%
\begin{align}
\sigma_{l m 0} \equiv {\rm Re}(\sigma_{l m 0}) + i \, {\rm Im}(\sigma_{l m 0})
= \omega_{l m 0} - \frac{i}{\tau_{l m 0}}\,,
\end{align}
%
are the complex frequencies of the fundamental ($0$-th overtone) QNMs of the remnant BH.
%
%\ag{Going back to the convention established above to use the index $k$ to now denote the $(l, m,0)$
%mode,  the frequencies $\omega_k$ and damping times $\tau_k$ can be read off from the
%real and imaginary parts of $\sigma_k$ respectively.}
%
The functions $\tilde{A}_{l m}(t)$ and $\tilde{\phi}_{l m}(t)$ are defined
in Ref.~\cite{Bohe:2016gbl,Cotesta:2018fcv}.
